---
title: "script for the analysis of Pomatias"
author: "Maxime Dahirel"
date: "13/02/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The present R Markdown file contains the analysis script for the manuscript entitled "Increased population density depresses activity but does not influence dispersal in the snail Pomatias elegans" (authors: Maxime Dahirel, Loic Menut, and Armelle Ansart), along with narrative comments to help the reader. 

(A "classical" R script version is also available in the same folder, for readers who'd rather use that. That version is much less commented, but both run the same analyses)


# Introduction

The aims of this study were to:

- determine whetehr dispersal and activity are density-dependent in Pomatias elegans
- determine whether these two traits were linked
-determine whether there was a link between behaviour and morphological defence, using (relative) shell weight as a proxy for the latter

To do so, we exposed snails to different densities and examined their dispersal and activity responses (binary traits); we then weighted their shells. We analysed these data using a multivariate mixed model approach.

(see manuscript for details)


# Raw data description and wrangling

First, let's load the packages we'll need:

```{r load-packages}
library(arm)
library(matrixStats)
library(tidyverse)
library(rstan)
library(tidybayes)
library(brms)
library(patchwork)
rstan_options(auto_write = TRUE)
options(mc.cores = 2)
```

Now we load the three raw data files:
```{r load-raw-datafiles}
## Load raw dataset
data <- read.table("./dataset_pomatias.txt", header = TRUE)
morpho <- read.table("./dataset_morpho_pomatias.txt", header = TRUE)
data_discrimin <- read.table("./discrimin_pomatias.txt", header = TRUE)
```

METADATA TO ADD HERE

For our analysis, we're going to need to (a) combine the "data" and "morpho" datasets, (b) create a bunch a new variables from the original ones, and (c) convert the dataset into a "long" format where both shell mass measures are in the same column, as well as both behavioural measures (the latter is not strictly needed, but it makes fitting a "latent" behavioural variable way easier).

So let's do that:

```{r data-wrangling1}
## creation of new variables
data<-as_tibble(data) %>% 
  mutate(BOX = factor(interaction(Date, Box_code)),
         Disp2 = -0.5 + as.numeric(Disp == "yes"), ## centered dummy variable ##for supplementary material only
         is.female = -0.5 + as.numeric(Sex_true == "F"), ## centered dummy variable
         scale_Density = scale(Density),
         uniqueID = as.numeric(str_split_fixed(ID,"_",3)[,2])
  ) %>% 
  left_join(morpho) %>% 
  mutate(behave_Disp=as.numeric(Disp=="yes"),behave_Active=Active) %>% 
  dplyr::select(Date,BOX,uniqueID,ID,
         Density,scale_Density,
         Sex_predicted,Sex_dissection=Sex_true,is.female,###rename Sex variable to something more "accurate"
         Height,Shell_width=Diametre,ApertureH=PeristomeH,  ###rename Diametre and PeristomeH variables to something more accurate
         Shell_Mass1,Shell_Mass2,
         Disp2,behave_Active,behave_Disp,Time_sec)

data1<-data %>% 
   dplyr::select(-c(behave_Active, behave_Disp)) %>% 
  pivot_longer(names_to="mass_measurement",
               values_to="Shell_Mass",
               cols=contains("Shell_Mass")) %>% 
  mutate(measurement_order=as.numeric(mass_measurement=="Shell_Mass2")+1)
data2<-data %>% 
   dplyr::select(-c(Shell_Mass1,Shell_Mass2)) %>% 
   pivot_longer(names_to="which_behaviour",
                values_to="behave",
                cols=contains("behave")) %>% 
  mutate(measurement_order=as.numeric(which_behaviour=="behave_Active")+1)


data<-left_join(data1,data2) %>% 
  mutate(which_behaviour=fct_recode(which_behaviour,Active="behave_Active",Disp="behave_Disp"))
```

```{r data-wrangling2}
data %<>% 
  mutate(scale_shell=scale(Shell_Mass),
       scale_height=scale(Height),
       scale_logshellmass=scale(log(Shell_Mass)),
       scale_logheight=scale(log(Height))) #%>% 
  #mutate(scale_logshellmass_NAfilled=replace_na(scale_logshellmass,0), ###needed for the subset() approach
  #     is.shell.valid=as.numeric(is.na(Shell_Mass)==FALSE))  ### the is.shell.valid variable ensures the filled NAs are ignored during fitting for the subset() approach
```

... and *Voilà*! We have our final dataset! But what's in it, you may ask? Let's look:

# Final data table description


# Main model

```{r main-model}
####main model
prior=c(
  set_prior("normal(0,1)",class="Intercept",resp=c("scalelogshellmass")),
  set_prior("normal(0,1)",class="b",resp=c("behave","scalelogshellmass")),
  set_prior("normal(0,1.5)",class="b",resp="behave",coef=c("which_behaviourActive","which_behaviourDisp")),
  set_prior("normal(0,1)",class="sd",resp=c("behave","scalelogshellmass")),
  set_prior("normal(0,1)",class="sigma",resp="scalelogshellmass"), #to relax?? hist(exp(rexp(1000,10)*0.19))
  set_prior("lkj(2)",class="cor")
)

bf_behave<-bf(behave~0+which_behaviour+which_behaviour:(scale_Density+scale_logheight+is.female)+
                (0+which_behaviour|BOX)+(1|q|ID),family=bernoulli)

bf_shell_miss<-bf(scale_logshellmass|mi() ~ scale_logheight + is.female+(1|q|ID))

##an alternative would be to use the subset approach, essentially ignoring the NA values, but only for the response with NA
## that is, no deletion of entire rows
### leads to sensibly the same results in our case
#bf_shell_subset<-bf(scale_logshellmass_NAfilled|subset(is.shell.valid) ~ scale_logheight + is.female+(1|q|ID))

mod=brm(mvbf(bf_behave+bf_shell_miss),data=data,prior=prior,iter=250, warmup=50,
        control=list(adapt_delta=0.95,max_treedepth=15),chains=4,seed=42) 

```


```{r some-posterior-checking}
```


# Plotting

```{r fig1}
newdata=data %>%
  select(which_behaviour) %>% 
  distinct() %>%
  expand_grid(BOX=data$BOX[1],ID=data$ID[1],is.female=0,scale_logheight=0,Density=c(1:30)/1) %>% 
  mutate(scale_Density=(Density-attr(data$scale_Density,"scaled:center"))/attr(data$scale_Density,"scaled:scale")) %>% 
  add_fitted_draws(mod,resp="behave",re_formula = NA)


p1<-newdata %>% 
  filter(which_behaviour=="Disp") %>% 
  ggplot()+
  stat_lineribbon(aes(x=Density,y=.value),.width=0.95,fill="grey90") + 
  geom_jitter(data=data %>% filter(which_behaviour=="Disp") %>% group_by(Density,BOX) %>% summarise(behave=mean(behave)),aes(x=Density,y=behave),col="grey60")+
  geom_point(data=data %>% filter(which_behaviour=="Disp") %>% group_by(Density) %>% summarise(behave=mean(behave)),aes(x=Density,y=behave),cex=3)+
  scale_x_continuous("")+
  scale_y_continuous("Dispersal rate")+
  theme_bw()+theme(legend.position = "none")

p2<-newdata %>% 
  filter(which_behaviour=="Active") %>% 
  ggplot()+
  stat_lineribbon(aes(x=Density,y=.value),.width=0.95,fill="grey90") + 
  geom_jitter(data=data %>% filter(which_behaviour=="Active") %>% group_by(Density,BOX) %>% summarise(behave=mean(behave)),aes(x=Density,y=behave),col="grey60")+
  geom_point(data=data %>% filter(which_behaviour=="Active") %>% group_by(Density) %>% summarise(behave=mean(behave)),aes(x=Density,y=behave),cex=3)+
  scale_x_continuous("Density (snails per box)")+
  scale_y_continuous("Probability of activity")+
  theme_bw()+theme(legend.position = "none")

(p1/p2)+plot_annotation(tag_levels="A")

```

```{r fig2}

newdata<-data %>% select(which_behaviour,behave,ID) %>% distinct()
newdata<-as_tibble(ranef(mod,summary = FALSE)$ID[,,"behave_Intercept"]) %>% 
  mutate(
    post_disp=select(.,all_of(subset(newdata$ID,newdata$which_behaviour=="Disp" & newdata$behave==1))) %>% rowMeans(),
    post_res=select(.,all_of(subset(newdata$ID,newdata$which_behaviour=="Disp" & newdata$behave==0))) %>% rowMeans(),
    post_act=select(.,all_of(subset(newdata$ID,newdata$which_behaviour=="Active" & newdata$behave==1))) %>% rowMeans(),
    post_still=select(.,all_of(subset(newdata$ID,newdata$which_behaviour=="Active" & newdata$behave==0))) %>% rowMeans(),
  ) %>% 
  cbind(posterior_samples(mod) %>% 
          select(meanlogitActive=b_behave_which_behaviourActive,meanlogitDisp=b_behave_which_behaviourDisp)) %>% 
  mutate(post_disp=post_disp+meanlogitActive,post_res=post_res+meanlogitActive,
         post_act=post_act+meanlogitDisp,post_still=post_still+meanlogitDisp) %>% 
  select(post_disp,post_res,post_act,post_still)

behave_wide<-data %>% 
  select(which_behaviour,behave,ID) %>% 
  pivot_wider(names_from=which_behaviour,values_from=behave)

p4 <- newdata %>% 
  select(post_act,post_still) %>% 
  pivot_longer(everything(),names_to="Active") %>%
  mutate(Active=fct_recode(Active,`0`="post_still",`1`="post_act")) %>% 
  ggplot()+
  geom_col(data=behave_wide %>% 
             group_by(Active) %>% 
             summarise(Disp=mean(Disp)),aes(x=factor(Active),y=Disp),col="black",fill="white")+
  geom_eye(aes(x=Active,y=invlogit(value)),.width=c(0,0.95))+
  scale_x_discrete("Active post dispersal?",labels=c("no","yes"))+
  scale_y_continuous("Dispersal rate",limits=c(0,1))+
  theme_bw()

(newdata$post_act-newdata$post_still) %>% median_hdi()
(newdata$post_disp-newdata$post_res) %>% median_hdi()

p5 <- newdata %>% 
  select(post_disp,post_res) %>% 
  pivot_longer(everything(),names_to="Disp") %>%
  mutate(Disp=fct_recode(Disp,`0`="post_res",`1`="post_disp")) %>% 
  ggplot()+
  geom_col(data=behave_wide %>% 
             group_by(Disp) %>% 
             summarise(Active=mean(Active==1)),aes(x=factor(Disp),y=Active),col="black",fill="white")+
  geom_eye(aes(x=Disp,y=invlogit(value)),.width=c(0,0.95))+
  scale_x_discrete("Dispersal status",labels=c("Resident","Disperser"))+
  scale_y_continuous("Activity probability",limits=c(0,1))+
  theme_bw()

p5+p4
```

```{r fig3}
newdata<-data %>% select(which_behaviour,behave,ID) %>% distinct()
newdata<-as_tibble(ranef(mod,summary = FALSE)$ID[,,"scalelogshellmass_Intercept"]) %>% 
  mutate(
    post_disp=select(.,all_of(subset(newdata$ID,newdata$which_behaviour=="Disp" & newdata$behave==1))) %>% rowMeans(na.rm=TRUE),
    post_res=select(.,all_of(subset(newdata$ID,newdata$which_behaviour=="Disp" & newdata$behave==0))) %>% rowMeans(na.rm=TRUE),
    post_act=select(.,all_of(subset(newdata$ID,newdata$which_behaviour=="Active" & newdata$behave==1))) %>% rowMeans(na.rm=TRUE),
    post_still=select(.,all_of(subset(newdata$ID,newdata$which_behaviour=="Active" & newdata$behave==0))) %>% rowMeans(na.rm=TRUE),
  ) %>% 
  mutate(post_disp=post_disp,post_res=post_res,
         post_act=post_act,post_still=post_still) %>% 
  select(post_disp,post_res,post_act,post_still)


p6 <- newdata %>% 
  select(post_act,post_still) %>% 
  pivot_longer(everything(),names_to="Active") %>%
  mutate(Active=fct_recode(Active,`0`="post_still",`1`="post_act")) %>% 
  ggplot()+
  geom_eye(aes(x=fct_rev(Active),y=exp(value*attr(data$scale_logshellmass,"scaled:scale"))),.width=c(0,0.95))+
  scale_x_discrete("Active post dispersal?",labels=c("no","yes"))+
  scale_y_continuous("",limits=c(0.95,1.05))+
  geom_hline(yintercept=1,lty=2)+
  theme_bw()

exp((newdata$post_act-newdata$post_still)*attr(data$scale_logshellmass,"scaled:scale")) %>% median_hdi()
exp((newdata$post_disp-newdata$post_res)*attr(data$scale_logshellmass,"scaled:scale")) %>% median_hdi()

p7 <- newdata %>% 
  select(post_disp,post_res) %>% 
  pivot_longer(everything(),names_to="Disp") %>%
  mutate(Disp=fct_recode(Disp,`0`="post_res",`1`="post_disp")) %>% 
  ggplot()+
  geom_eye(aes(x=fct_rev(Disp),y=exp(value*attr(data$scale_logshellmass,"scaled:scale"))),.width=c(0,0.95))+
  scale_x_discrete("Dispersal status",labels=c("Resident","Disperser"))+
  scale_y_continuous("Relative Shell mass",limits=c(0.95,1.05))+
  geom_hline(yintercept=1,lty=2)+
  theme_bw()

(p7+p6)
```

# Supplementary Materials

```{r supplementary-models}

####models for supplementary

mod_S1a <- data %>% filter(mass_measurement=="Shell_Mass1") %>% 
  group_by(BOX) %>% 
  summarise(Density=mean(Density),Nfemales=sum(Sex_dissection=="F")) %>% 
  ungroup() %>% 
  brm(Nfemales|trials(Density)~1,family=binomial,
      prior=c(set_prior("normal(0,1.5)",class="Intercept")),data=., seed = 42
  )

mod_S1b <- data %>% filter(mass_measurement=="Shell_Mass1") %>% 
  group_by(BOX) %>% 
  summarise(Density=mean(Density),Nfemales=sum(Sex_dissection=="F")) %>% 
  ungroup() %>% 
  brm(Nfemales|trials(Density)~0+factor(Density),family=binomial,
      prior=c(set_prior("normal(0,1.5)",class="b")),data=., seed = 42
  )

mod_S3 <- brm(behave~scale_Density+scale_logheight+is.female+Disp2+(1|BOX),family=bernoulli,
             prior=c(set_prior("normal(0,1.5)",class="Intercept"),
                     set_prior("normal(0,1)",class="b"),
                     set_prior("normal(0,1)",class="sd")),data=subset(data,which_behaviour=="Active"), seed = 42
)
```
